**Sequencing 16S rRNA gene fragments using the PacBio SMRT DNA sequencing system**

### 2016-02-03 Schloss_PacBio16S_PeerJ_2016.Rmd
Build Figure 1

***The PacBio error profile and a basic sequence curation procedure.***
To build a sequence curation pipeline, we first needed to characterize the error rate associated with sequencing the 16S rRNA gene. We observed an average sequencing error rate of `r format(error_profile$overall_error, digits=2, nsmall=2)`%. Insertions, deletions, and substitutions accounted for `r format(error_profile$ins_rate, digits=3, nsmall=1)`, `r format(error_profile$del_rate, digits=3, nsmall=1)`, and `r format(error_profile$subs_rate, digits=3, nsmall=1)`% of the errors, respectively. The substitution errors were equally likely and all four bases were equally likely to cause insertion errors. Interestingly, guanines (`r format(error_profile$del_bias$dG, digits=3, nsmall=1)`%) and adenines (`r format(error_profile$del_bias$dA, digits=3, nsmall=1)`%) were more likely to be deleted than cytosines (`r error_profile$del_bias$dC, digits=3, nsmall=1)`%) or thymidines (`r error_profile$del_bias$dT, digits=2, nsmall=1)`%). The PacBio quality values varied between `r paste(error_profile$qual_range, collapse=' and ')`. Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(error_profile$percent_over_93$m, digits=3, nsmall=1)`%), substitutions (`r format(error_profile$percent_over_93$s, digits=3, nsmall=1)`%), and insertions (`r format(error_profile$percent_over_93$i, digits=3, nsmall=1)`%). Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(error_profile$qual_error_corr, digits=2, nsmall=2)`; Figure 1A).

@@

[revise - Fig 1A plot predicted error rate vs observed error rate]
**Figure 1. Summary of errors in data generated using PacBio sequencing platform
to sequence various regions within the 16S rRNA gene.** Quality scores varied
with error types (A). The sequencing error rate of the amplified gene fragments
increased with mismatches to the barcodes and primers (B). The sequencing error
rate declined with increasd sequencing coverage; however, increasing the
sequencing depth beyond 10-fold coverage had no meaningful effect on the
sequencing error rate (C).

@@

### Figure 1
```{r Figure_1, fig.width=6.5, fig.height=3.0}
layout(matrix(c(1,2,3), nrow=1), heights=1, widths=c(0.75,1,1))

#Panel A depicting range of quality scores for different errors
par(mar=c(6, 5, 0.5, 0.5))
plot(1, xlim=c(0.5, 4.5), ylim=c(0,75), type="n", yaxt="n", xaxt="n", xlab="", ylab="")
polygon(x=c(0.75, 1.25, 1.25, 0.75), y=c(m.quant["25%"], m.quant["25%"],
											 m.quant["75%"], m.quant["75%"]))
polygon(x=c(1.75, 2.25, 2.25, 1.75), y=c(s.quant["25%"], s.quant["25%"],
											 s.quant["75%"], s.quant["75%"]))
polygon(x=c(2.75, 3.25, 3.25, 2.75), y=c(i.quant["25%"], i.quant["25%"],
											 i.quant["75%"], i.quant["75%"]))
polygon(x=c(3.75, 4.25, 4.25, 3.75), y=c(a.quant["25%"], a.quant["25%"],
											 a.quant["75%"], a.quant["75%"]))

segments(x0=seq(0.75, 3.75, 1), x1=seq(1.25,4.25,1), y0=c(m.quant["50%"],
											s.quant["50%"], i.quant["50%"], a.quant["50%"]), lwd=4)

arrows(x0=1:4, y0=c(m.quant["75%"], s.quant["75%"], i.quant["75%"],
					a.quant["75%"]), y1=c(m.quant["97.5%"], s.quant["97.5%"],
					i.quant["97.5%"], a.quant["97.5%"]), angle=90, length=0.05)

arrows(x0=1:4, y0=c(m.quant["25%"], s.quant["25%"], i.quant["25%"],
					a.quant["25%"]), y1=c(m.quant["2.5%"], s.quant["2.5%"],
					i.quant["2.5%"], a.quant["2.5%"]), angle=90, length=0.05)

axis(1, at=c(1,2,3,4), label=c("Matches", "Substitutions", "Insertions", "Ambiguous"), las=2, cex.axis=1)
axis(2, las=2, at=c(0,25,50,75), cex.axis=1.5)
title(ylab="Quality score", cex.lab=1.5)
mtext(side=2, line=2.5, at=75, text="A", cex=2, font=2, las=2)

#Panel B depicting relationship between errors in barcodes/primers and the
#rest of the sequence.
par(mar=c(5,5,0.5,0.5))
plot(err.bcprimer$x[1:6,1]~err.bcprimer$Group.1[1:6], pch=19,
		 ylab="Error rate (%)", xlab="", axes=F, ylim=c(0,5), cex.lab=1.5)
mtext(side=1, line=3.75, text="Total mismatches to\nbarcodes and primers",
			cex=0.9)
axis(1, cex.axis=1.5)
axis(2, las=2, cex.axis=1.5)
box()
mtext(side=2, line=2.5, at=5, text="B", cex=2, font=2, las=2)

#Panel C depicting relationship between sequencing coverage and error rate for
#the different regions that were sequence
plotLines <- function(folder){
	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
								 ".composite", sep=""), header=T,row.names=1)

	composite <- composite[composite$reason=="g",]
	error.by.depth <- aggregate(composite$error, by=list(composite$freq),
						function(x){c(mean=mean(x), quantile(x, prob=c(0.025, 0.975)), n=length(x))})

	#exclude any coverage values where we don't have at least 50 observations
	error.by.depth <- error.by.depth[error.by.depth$x[,"n"]>=50,]
	fold.coverage <- error.by.depth[, "Group.1"]
	error.by.depth <- error.by.depth$x

	mean <- error.by.depth[,"mean"]
	points(mean~fold.coverage, type="l", lwd=2, col=clrs[folder])
}

par(mar=c(5,2,0.5, 0.5))
plot(1, xlim=c(2,30), ylim=c(0,0.05), type="n", xlab="", ylab="", axes=F)
axis(2, at=seq(0,0.05, 0.01), label=FALSE, tick=T, las=2)
#axis(2, at=seq(0,0.05, 0.01), label=seq(0,5, 1), las=2)
axis(1, cex.axis=1.5)
lapply(regions, plotLines)
legend(x=18, y=0.05, legend=pretty_region[regions], lty=1, lwd=2, col=clrs[regions])
mtext(side=2, line=0.5, at=0.05, text="C", cex=2, font=2, las=2)
title(xlab="Coverage", ylab="", cex.lab=1.5)
box()

>>>>

***The PacBio error profile and a basic sequence curation procedure.***
To build a sequence curation pipeline, we first needed to characterize the error rate associated with sequencing the 16S rRNA gene. We observed an average sequencing error rate of `r format(error_profile$overall_error, digits=2, nsmall=2)`%. Insertions, deletions, and substitutions accounted for `r format(error_profile$ins_rate, digits=3, nsmall=1)`, `r format(error_profile$del_rate, digits=3, nsmall=1)`, and `r format(error_profile$subs_rate, digits=3, nsmall=1)`% of the errors, respectively. The substitution errors were equally likely and all four bases were equally likely to cause insertion errors. Interestingly, guanines (`r format(error_profile$del_bias$dG, digits=3, nsmall=1)`%) and adenines (`r format(error_profile$del_bias$dA, digits=3, nsmall=1)`%) were more likely to be deleted than cytosines (`r error_profile$del_bias$dC, digits=3, nsmall=1)`%) or thymidines (`r error_profile$del_bias$dT, digits=2, nsmall=1)`%). The PacBio quality values varied between `r paste(error_profile$qual_range, collapse=' and ')`. Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(error_profile$percent_over_93$m, digits=3, nsmall=1)`%), substitutions (`r format(error_profile$percent_over_93$s, digits=3, nsmall=1)`%), and insertions (`r format(error_profile$percent_over_93$i, digits=3, nsmall=1)`%). Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(error_profile$pred_error_corr, digits=2, nsmall=2)`; Figure 1A).

@@

**Figure 1. Summary of errors in data generated using PacBio sequencing platform
to sequence various regions within the 16S rRNA gene.** The predicted error rate using PacBio's sequence analysis algorithm correlated well with the observed error rate (Pearson's R: `r format(error_profile$pred_error_corr, digits=2, nsmall=2)`; A). Because of the large number of sequences, we randomly selected 5% of the data to show in panel A. The sequencing error rate of the amplified gene fragments
increased with mismatches to the barcodes and primers (B). The sequencing error
rate declined with increased sequencing coverage; however, increasing the
sequencing depth beyond 10-fold coverage had no meaningful effect on the
sequencing error rate (C).

@@
